

# Optimize your 401-k

* [Abstract](#Abstract)
* [Load Libraries](#Load-Libraries)
* [Download Monthly Quotes](#Download-Fund-Ticker-Quotes)
* [Compute Monthly Returns](#Compute-Simple-Returns)
* [Clustering](#Clustering)
* [PCA: Visualizing Clusters](#PCA--Visualizing-Clusters)
* [Chart Cumulative Returns](#Chart-Cumulative-Returns)
* [Qualitative Evaluation of Clusters](#Qualitative-Evaluation-of-Clusters)
* [Plot Selected Funds on PCA Chart](#Plot-Selected-Funds-on-PCA-Chart)
* [Mean-Var Analysis of Selected Funds](#Mean-Var-Analysis-of-Selected-Funds)
* [Portfolio Analysis of Selected Funds](#Portfolio-Analysis-of-Selected-Funds)
* [Back-test Selected Portfolio](#Back-test-Selected-Portfolio)
* [Null Funds](#Null-Funds)
* [Include Null Funds](#Include-Null-Funds)
* [Future](#Future)


## <a name="Abstract"></a>Abstract

Allocating money across the various investment funds available in your 401-k 
account can be challenging.  Some 401-k managers offer hundreds of funds to choose from.
The large number of options makes it difficult to construct an optimal investment portfolio. 

This article tackles the problem using the following strategy:

1. Reduce the number of options by sorting the funds into a set of
"clusters" based on similarity between the funds' monthly rate-of-returns
(i.e. funds with highly correlated monthly returns are sorted together in the same cluster).
2. Select a single "representative fund" from each cluster.  
3. Combine the representative funds into an optimal investment portfolio.

Step 1 involved using K-means clustering to cluster funds based on the 
past 5 years of monthly returns.  The optimal number of clusters (10) was 
decided using multiple cluster evaluation metrics.  The clustering
algorithm performed well, in that funds sorted themselves together as an 
investor might expect them to (large-cap stocks clustered together, high-yield 
bonds clustered together, real estate funds clustered together, etc).

Step 2 involved using mean-variance analysis to select an optimally
performing fund as the "representative fund" for each cluster.  In most
cases the representative fund was easy to choose as it dominated the 
mean-variance analysis (the fund produced the best mean return at nearly the 
lowest volatility).  In other cases, where no fund clearly dominated
the cluster, the representative fund was selected by trading off return 
and volatility (in a somewhat arbitrary manner according to my own risk appetite).

Step 3 involved using standard mean-variance portfolio analysis to determine
an optimal allocation strategy across the representative funds.

Note that this strategy generalizes to any set of assets, not just 401-k mutual funds.

**Caveat emptor**: It goes without saying that correlation does not equal causation,
that past results do not guarantee future performance.  This article does not constitute 
investment advice.  Trade at your own risk.


## <a name="Load-Libraries"></a>Load Libraries

```{r}
    
    library(tseries)                # get.hist.quote
    library(zoo)                    # coredata
    library(quadprog)               # solve.QP
    library(PerformanceAnalytics)   # chart.*
    library(dplyr)                  # mutate, arrange, filter

    source("mean-var.R")
    source("my401k-functions.R")
    source("my401k-etl.R")
    source("my401k-clustering.R")

    options(width=150)              # R results display width
    options(scipen=7)               # favor fixed notation over scientific notation

```


## <a name="Download-Fund-Ticker-Quotes"></a>Download Monthly Quotes


```{r}

    funds.df <- My401kEtl$loadFundList()

    #
    # zoo time-series object
    # rows: monthly quotes
    # cols: funds
    #
    quotes.zoo <- My401kEtl$loadAllTickerQuotes(funds.df)

    #
    # print the number of monthly quotes for each fund
    #
    quote.counts <- apply( !is.na(quotes.zoo),2,sum)
    sort(quote.counts, decreasing=T)

```


## <a name="Compute-Simple-Returns"></a>Compute Monthly Returns

```{r}

    simple.returns.zoo <- My401k$computeSimpleReturns(quotes.zoo)

    simple.returns.fees.zoo <- My401k$computeSimpleReturnsWithFees(simple.returns.zoo,
                                                                   funds.df)

    simple.returns.stats.df <- My401k$computeSimpleReturnsStats(simple.returns.zoo)
    simple.returns.fees.stats.df <- My401k$computeSimpleReturnsStats(simple.returns.fees.zoo)

    # fundsx.df <- merge(funds.df, simple.returns.stats.df, by=0, all.x=T)

    #
    # Extract the core matrix data
    # transpose: each row is a fund, each col is a monthly return
    #
    simple.returns.mat <- t( coredata(simple.returns.zoo) )

```


## <a name="Clustering"></a>Clustering


```{r}


    #
    # list of objects returned by kmeans()
    # one for each number-of-clusters (4:14)
    #
    km.list <- My401kClustering$generateKmeansResults(simple.returns.mat, 4:14)

    #
    # Various measures to evaluate clustering results
    # Some measures (DB, SD) suggest 9 clusters
    # Other measures (R^2, Hartigan) suggest 11 clusters
    # Go with 10?
    #
    km.eval <- My401kClustering$evaluateKmeansResults(km.list, simple.returns.mat)
    data.frame(km.eval)

    km <- km.list[[7]]      # 10 clusters

```

## <a name="PCA--Visualizing-Clusters"></a>PCA: Visualizing Clusters


```{r}

    pca <- prcomp(simple.returns.mat, 
                  center = T)

    #
    # 90% of variance is explained by first 9 principal components.
    # This also suggests using ~9 clusters, tho I can't explain why.
    #
    summary(pca)$importance[,1:11]

    #
    # pca$x is the projection of simple.returns.mat (centered) onto the principal components.
    # Use the first two principal components to project the high-dimensional data onto a plane
    # for visualization.
    # 
    proj.2d <- pca$x[,1:2]
    plot(proj.2d, 
         col=km$cluster, 
         pch=20,
         xlim=c(-0.3,0.2), 
         ylim=c(-0.12,0.35))
    title("Projection of monthly return data\non first 2 PC's")

    #
    # legend: cluster id (# of funds in cluster)
    #
    legend.text = sapply(1:10, function(i) { paste(toString(i), " (", toString(km$size[i]), ")", sep="") })
    legend(x="topleft", legend=legend.text, col=1:10, pch=20)

    #
    # Include 3rd PC
    #
    proj.2d <- pca$x[,3:2]
    plot(proj.2d, 
         col=km$cluster, 
         pch=20, 
         xlim=c(-0.3,0.2), 
         ylim=c(-0.12,0.35))
    title("Projection of monthly return data\non 2nd and 3rd PC's")


```


## <a name="Chart-Cumulative-Returns"></a>Chart Cumulative Returns


```{r}


    #
    # list of vectors of ticker symbols 
    #
    cluster.tickers.list <- My401kClustering$getClusterTickersList( km ) 

    #
    # list of lists of funds
    #
    cluster.funds.list <- My401kClustering$getClusterFundsList( cluster.tickers.list, funds.df) 


    My401k$chartCumulativeReturns( simple.returns.fees.zoo,
                                   simple.returns.fees.stats.df,
                                   cluster.tickers.list,
                                   cluster.funds.list )


```

## <a name="Qualitative-Evaluation-of-Clusters"></a>Qualitative Evaluation of Clusters


* Cluster 1 
    * 20/20 foreign stock investments
    * Primarily large cap, Japan, Europe
    * A handful of small/mid cap, emerging markets, and real estate
    * Selected fund: FISMX: FID INTL SMALL CAP
        * Dominates mean-var
        * even after accounting for highest fees in the group
* Cluster 2
    * 6/6 foreign stock investments
    * Primarily emerging markets
    * Selected fund: FHKCX: FID CHINA REGION
        * Highest mean, also highest variance
* Cluster 3
    * 3/3 Real Estate funds
    * 2 domestic, 1 global
    * Selected fund: FRESX: FID REAL ESTATE INVS
        * mid-point on both mean and var
* Cluster 4
    * 36/39 bond funds, 2 blended funds, 1 real estate
    * Mostly intermediate-term govt bonds
    * low volatility
    * selected fund: VWIAX: VANG WELLESLEY ADM
        * blended fund
    * FRIFX: FID REAL ESTATE INC - out of place?
    * PFORX: PIM FOR BD USHG I - bond fund, more representative of the cluster
* Cluster 5
    * 15/15 small and mid-cap domestic stock funds
    * selected fund: VSEQX: VANG STRATEGIC EQ
        * dominates mean-var
        * with relatively low fees
* Cluster 6
    * 6/6 long-term domestic bonds
    * selected fund: VWETX   VANG LT INV GR ADM
        * dominates mean-var
* Cluster 7
    * 19 high-yield bonds, emerging markets, commodities, and aggressive blends
    * selected fund: FAGIX FID CAPITAL & INCOME
        * dominates mean-var
* Cluster 8
    * IBM only
* Cluster 9
    * 45 mostly large-cap stocks
    * selected fund: VGIAX: VANG GRTH & INC ADM
        * 1/6 reduction in variance for only 1/13 reduction in return
* Cluster 10
    * 15 mostly large-cap stocks
    * similar to cluster 9
    * selected fund:  RGAGX AF GRTH FUND AMER R6
        * 1/4 reduction in variance for only 1/13 reduction in return


## <a name="Plot-Selected-Funds-on-PCA-Chart"></a>Plot Selected Funds on PCA Chart

```{r}

    selected.funds.vec <- c( "FISMX",
                             "FHKCX",
                             "FRESX",
                             "VWIAX",
                             "FRIFX",
                             "PFORX",
                             "VSEQX",
                             "VWETX",
                             "FAGIX",
                             "VGIAX",
                             "RGAGX")

    #
    # Highlight selected funds on PCA chart
    #
    proj.2d <- pca$x[,1:2]
    plot(proj.2d, 
         col=km$cluster, 
         pch=20,
         xlim=c(-0.3,0.2), 
         ylim=c(-0.12,0.35))
    title("Projection of monthly return data\non first 2 PC's")

    #
    # legend: cluster id (# of funds in cluster)
    #
    legend.text = sapply(1:10, function(i) { paste(toString(i), " (", toString(km$size[i]), ")", sep="") })
    legend(x="topleft", legend=legend.text, col=1:10, pch=20)

    #
    # label selected funds
    #
    proj.2d.selected.funds <- proj.2d[ selected.funds.vec, ]
    text(x=proj.2d.selected.funds[,1],
         y=proj.2d.selected.funds[,2],
         labels=selected.funds.vec,
         pos=4,
         cex=0.75)

    #
    # Include 3rd PC
    #
    proj.2d <- pca$x[,3:2]
    plot(proj.2d, 
         col=km$cluster, 
         pch=20, 
         xlim=c(-0.3,0.2), 
         ylim=c(-0.12,0.35))
    title("Projection of monthly return data\non 2nd and 3rd PC's")

    #
    # label selected funds
    #
    proj.2d.selected.funds <- proj.2d[ selected.funds.vec, ]
    text(x=proj.2d.selected.funds[,1],
         y=proj.2d.selected.funds[,2],
         labels=selected.funds.vec,
         pos=4,
         cex=0.75)

    #
    # Remove FRIFX and VWIAX (not representative of cluster 4)
    #
    selected.funds.vec <- c( "FISMX",
                             "FHKCX",
                             "FRESX",
                             "PFORX",
                             "VSEQX",
                             "VWETX",
                             "FAGIX",
                             "VGIAX",
                             "RGAGX")


```


## <a name="Mean-Var-Analysis-of-Selected-Funds"></a>Mean-Var Analysis of Selected Funds


```{r}

    #
    # Filter selected funds from the returns data
    #
    selected.returns.zoo <- simple.returns.fees.zoo[, selected.funds.vec]
    selected.returns.stats.df <- simple.returns.fees.stats.df[ selected.funds.vec, ]
    selected.funds.df <- funds.df[ selected.funds.vec, ]

    My401k$generateReturnsCharts( selected.returns.zoo,
                                  selected.returns.stats.df,
                                  selected.funds.df )
                                   
```

## <a name="Portfolio-Analysis-of-Selected-Funds"></a>Portfolio Analysis of Selected Funds


Note: current risk-free rate (US10YR=2.45%, mid-Feb 2017) found
[here](https://www.treasury.gov/resource-center/data-chart-center/interest-rates/Pages/TextView.aspx?data=yield).


```{r}

    #
    # build the efficient frontier (no shorting!).
    # First calc the GMV portfolio.
    # Then generate EFF portfolios from the GMV mean/var to max(asset-mean)
    #
    portfolio <- MeanVar$initPortfolio(asset.vec=as.character(selected.returns.stats.df$ticker),
                                       mu.vec=selected.returns.stats.df$mean,
                                       stdev.vec=selected.returns.stats.df$sd,
                                       Sigma.mat=coredata( cov( selected.returns.zoo )),
                                       tan.rf.rate=(2.45^(1/12) - 1)/100 )      # monthly rate assuming monthly compounding

    portfolio <- MeanVar$buildPortfolio.noShort( portfolio )
    portfolio

    #
    # Generate the mean-var plot
    #
    MeanVar$plotPortfolio( portfolio )

    #
    # Plot and list asset weights
    #
    MeanVar$plotEffAssetWeights( portfolio, main="Asset weights of efficient portfolios" )
    MeanVar$printEffPortfolios( portfolio )



```

## <a name="Back-test-Selected-Portfolio"></a>Back-test Selected Portfolio


```{r}

    #
    # selected efficient portfolio #10
    #
    portfolio$w.vec <- portfolio$eff.w.mat[,10]

    #
    # Generate monthly portfolio returns.
    # 
    # selected.returns.zoo contains the asset returns.
    # portfolio$w.vec contains the asset weights.
    #  (n x k) * (k x 1) -> (n x 1)
    # dot each row (row=monthly return, column=asset)
    # with the weight vector
    # to get the portfolio's monthly returns
    #
    port.returns.zoo <- zoo( selected.returns.zoo %*% portfolio$w.vec,
                             order.by=index(selected.returns.zoo))
    colnames(port.returns.zoo) <- c("PORT")

    #
    # Merge portfolio returns with individual asset returns 
    #
    port.returns.zoo <- merge(selected.returns.zoo, port.returns.zoo)
    port.returns.stats.df <- My401k$computeSimpleReturnsStats( port.returns.zoo )   # dup effort here but oh well

    #
    # Plot cum returns and mean var of the portfolio + individual assets
    #
    My401k$plotCumReturnsAndMeanVar( port.returns.zoo,
                                     port.returns.stats.df,
                                     lty=c(rep(1,ncol(selected.returns.zoo)), 2) )

    #
    # Plot histogram of just port returns
    #
    hist(port.returns.zoo$PORT, 
         breaks=20,
         col="lightblue")

```

## <a name="Null-Funds"></a>Null Funds

```{r}

    #
    # 4 funds always have 0 allocation
    # I want to include them anyway, for diversification
    # How to do?
    # Steal equal % from all other allocations
    # divvy among the 4 - but how? mean-var analysis on just the 4?
    #
    # 
    null.funds.vec <- c("FISMX", "FHKCX", "FRESX", "FAGIX")

    #
    # Filter selected funds from the returns data
    #
    null.returns.zoo <- simple.returns.fees.zoo[, null.funds.vec]
    null.returns.stats.df <- simple.returns.fees.stats.df[ null.funds.vec, ]
    null.funds.df <- funds.df[ null.funds.vec, ]

    My401k$generateReturnsCharts( null.returns.zoo,
                                  null.returns.stats.df,
                                  null.funds.df )

    #
    # build the efficient frontier (no shorting!).
    #
    null.portfolio <- MeanVar$initPortfolio(asset.vec=as.character(null.returns.stats.df$ticker),
                                            mu.vec=null.returns.stats.df$mean,
                                            stdev.vec=null.returns.stats.df$sd,
                                            Sigma.mat=coredata( cov( null.returns.zoo )),
                                            tan.rf.rate=(2.45^(1/12) - 1)/100 ) 

    null.portfolio <- MeanVar$buildPortfolio.noShort( null.portfolio )
    null.portfolio

    #
    # Generate the mean-var plot
    #
    MeanVar$plotPortfolio( null.portfolio )

    #
    # Plot and list asset weights
    #
    MeanVar$plotEffAssetWeights( null.portfolio, main="Asset weights of efficient portfolios" )
    MeanVar$printEffPortfolios( null.portfolio )


```

## <a name="Include-Null-Funds"></a>Include Null Funds

```{r}

    #
    # Select "null portfolio" #10
    #
    portfolio$w.vec <- portfolio$eff.w.mat[,10]
    null.portfolio$w.vec <- null.portfolio$eff.w.mat[,10]

    #
    # Construct various combinations of optimal portfolio and null portfolio:
    #   100% optimal, 0% null
    #   90% optimal, 10% null
    #   80% optimal, 20% null
    #   70% optimal, 30% null
    #   60% optimal, 40% null
    #   50% optimal, 50% null
    #
    w0.vec <- portfolio$w.vec

    w10.vec <- portfolio$w.vec * 0.9
    w10.vec[ names(null.portfolio$w.vec) ] <- null.portfolio$w.vec * 0.1

    w20.vec <- portfolio$w.vec * 0.8
    w20.vec[ names(null.portfolio$w.vec) ] <- null.portfolio$w.vec * 0.2

    w30.vec <- portfolio$w.vec * 0.7
    w30.vec[ names(null.portfolio$w.vec) ] <- null.portfolio$w.vec * 0.3

    w40.vec <- portfolio$w.vec * 0.6
    w40.vec[ names(null.portfolio$w.vec) ] <- null.portfolio$w.vec * 0.4

    w50.vec <- portfolio$w.vec * 0.5
    w50.vec[ names(null.portfolio$w.vec) ] <- null.portfolio$w.vec * 0.5

    w.df <- data.frame(PORT=w0.vec,
                       PORT10=w10.vec,
                       PORT20=w20.vec,
                       PORT30=w30.vec,
                       PORT40=w40.vec,
                       PORT50=w50.vec)

    colSums(w.df)   # sanity check - all cols should sum to 1

    #
    # Generate monthly portfolio returns.
    # 
    # selected.returns.zoo contains the asset returns (row=monthly return, column=asset)
    # w.df contains the asset weights (row=asset, column=port weights)
    #   (n x k) * (k x 5) -> (n x 5)    (row=monthly return, column=port
    # dot the fund monthly returns in each row in selected.returns.zoo
    # with the fund weights in each column of w.df
    # to get each portfolio's monthly returns
    #
    port.returns.zoo <- zoo( selected.returns.zoo %*% data.matrix(w.df),
                             order.by=index(selected.returns.zoo))

    #
    # Merge portfolio returns with individual asset returns 
    #
    port.returns.zoo <- merge(selected.returns.zoo, port.returns.zoo)
    port.returns.stats.df <- My401k$computeSimpleReturnsStats( port.returns.zoo )   # dup effort here but oh well

    #
    # Plot cum returns and mean var of the portfolio + individual assets
    #
    My401k$plotCumReturnsAndMeanVar( port.returns.zoo,
                                     port.returns.stats.df,
                                     lty=c(rep(1,ncol(selected.returns.zoo)), 
                                           rep(2,ncol(w.df)) ))

```

You can see why the "null.portfolio" funds were kept out of the optimal portfolio.
Adding them in not only pushes down the mean return, it also pushes up the volatility.
Nevertheless, there are good reasons why we should include them anyway:

1. Diversification!
2. Past returns don't necessarily predict future performance.

Better to be diversified than to bet all your money that the next 5 years will 
be just like the last 5 years.




## <a name="Future"></a>Future

* mean-cVaR analysis

    



